Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-14 paper.tex
Pass through ARG section

Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. The likelihood of an 
ARG is then a product over all these three classes of term.
In essence, computing the KYF likelihood requires
``replaying'' the  state of Hudson's simulation
algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
compute the exact configuration of lineages and distribution
of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
It therefore requires complete
detail about each of these events to be stored. 
Other approaches to defining the likelihood of an ARG under a coalescent-like 
model follow the same basic event-by-event 
strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
% [Are there others?  The bacterial stuff from Didelot etc seem to focus on
% a tree-by-tree likelihood, rather than global graph, so not clear it's the 
% same thing]
Computing the likelihood under the SMC rather than the Hudson coalescent 
simplifies bookkeeping somewhat,  % Dunno what I'm talking about here
but a complete event-by-event description of the 
input ARG is still required.

Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, such complete detail is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical \texttt{ms}
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The \texttt{msprime}
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

@@

\section*{ARG likelihoods}

In one of the first approaches to probilistic ARG inference that 
regarded the ARG itself as an output rather than a latent variable,
\cite{kuhner_maximum_2000} provided an explicit formula for 
the likelihood of a realised ARG under the coalescent with
recombination:
\[
\mathbb{P}[G|N, r] = \left(\frac{1}{2N}\right)^K
r^X \exp\left(-\sum \left( \frac{1}{2N} \binom{k_i}{2} + r\ell_i\right)
t_i\right),
\]
where [explain notation].
Under this KYF formulation, the likelihood is computed by taking a 
product of the probability of each realised event, and their 
inter-event waiting times. While it is straightforward to write 
this down, computation is not so easy, and in essence requires
In essence, computing the KYF likelihood requires
``replaying'' the  state of Hudson's simulation
algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
compute the exact configuration of lineages and distribution
of ancestral material at each time point~\citep{baumdicker_efficient_2021}.

Computing the likelihood of a realised ARG under the SMC simplifies the 
bookkeeping somewhat, because we no long have the possibility of 
``trapped'' ancestral material, and each lineage can be represented 
by a single interval representing their ancestral material. Nonetheless,
we still require complete knowledge of the graph to compute the 
required probablilities.

@@

% % % % % % % % % % % % % % % % % %
\subsection{ARGs} \label{par:recording}

Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
in the generic sense, it is a labeled graph structure that records
the inheritance of genetic material \citep{wong_general_2023}.
An ARG of the sort we consider describes the history of inheritance
of the genomic material of a focal set of individuals, or \emph{samples}.
Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
of the form ``$c$ inherited from $p$ on the genomic segment from $x$ to $y$'',
where $c$ and $p$ are (haploid) genomes, and $x$ and $y$ are genetic coordinates.
We record each such statement in an ``edge'',
often summarized as the tuple $(p,c,x,y)$.
By ``describes the history'',
we mean that each node in the ARG represents a specific (ancestral) genome,
and that the ARG specifies the genealogical tree describing how the samples are related
at each point on the genome.

@@

Crucial for the equivalence between both representations is the presence of
nodes that only have a single child along one or more local trees --
i.e., are \emph{locally unary}~\citep{wong_general_2023}.
Recombinant nodes (in blue) have a single child in all trees.
However, common ancestry nodes can be locally unary as well:
for instance, sample $d$ inherited a longer stretch of genome from ancestor $f$
than did sample $c$.
In the top plot, this is visible because the pink bar below node $f$
extends further to the left than does the blue bar,
while in the bottom plot, this is visible because node $f$ is parent to $d$
in the second tree, but not to node $c$, and so is unary.

@@

These locally unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify the ARG by removing all recombinant nodes
($A$, $B$, and $C$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombinant nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombinant nodes we no longer know the precise time of each recombination event,
but we do still have constraints on the time:
for instance, we know that $C$ happened somewhere between node $f$ and node $h$.
More generally, a recombinant node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG reconstruction methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently,
\argweaver{} \citep{rasmussen_genome-wide_2014},
\tsinfer{} \citep{kelleher_inferring_2019},
\kwarg{} \citep{ignatieva_kwarg_2021}, and
\texttt{SINGER} \citep{deng_robust_2024} estimate ARGs 
including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation, kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).

>>>>

Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. The likelihood of an 
ARG is then a product over all these three classes of term.
% In essence, computing the KYF likelihood requires
% ``replaying'' the  state of Hudson's simulation
% algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
% compute the exact configuration of lineages and distribution
% of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
% It therefore requires complete
% detail about each of these events to be stored. 
% Other approaches to defining the likelihood of an ARG under a coalescent-like 
% model follow the same basic event-by-event 
% strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
% % [Are there others?  The bacterial stuff from Didelot etc seem to focus on
% % a tree-by-tree likelihood, rather than global graph, so not clear it's the 
% % same thing]
% Computing the likelihood under the SMC rather than the Hudson coalescent 
% simplifies bookkeeping somewhat,  % Dunno what I'm talking about here
% but a complete event-by-event description of the 
% input ARG is still required.
Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, such complete detail is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical \texttt{ms}
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The \texttt{msprime}
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

@@

% \section*{ARG likelihoods}

% In one of the first approaches to probilistic ARG inference that 
% regarded the ARG itself as an output rather than a latent variable,
% \cite{kuhner_maximum_2000} provided an explicit formula for 
% the likelihood of a realised ARG under the coalescent with
% recombination:
% \[
% \mathbb{P}[G|N, r] = \left(\frac{1}{2N}\right)^K
% r^X \exp\left(-\sum \left( \frac{1}{2N} \binom{k_i}{2} + r\ell_i\right)
% t_i\right),
% \]
% where [explain notation].
% Under this KYF formulation, the likelihood is computed by taking a 
% product of the probability of each realised event, and their 
% inter-event waiting times. While it is straightforward to write 
% this down, computation is not so easy, and in essence requires
% ``replaying'' the  state of Hudson's simulation
% algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
% compute the exact configuration of lineages and distribution
% of ancestral material at each time point~\citep{baumdicker_efficient_2021}.

% % Computing the likelihood of a realised ARG under the SMC simplifies the 
% % bookkeeping somewhat, because we no long have the possibility of 
% % ``trapped'' ancestral material, and each lineage can be represented 
% % by a single interval representing their ancestral material. Nonetheless,
% % we still require complete knowledge of the graph to compute the 
% % required probablilities.

@@

% % % % % % % % % % % % % % % % % %
\subsection*{ARGs} \label{par:recording}
Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
in the generic sense, it is a labeled graph structure that records
the inheritance of genetic material \citep{wong_general_2023}.
An ARG of the sort we consider describes the history of inheritance
of the genomic material of a focal set of individuals, or \emph{samples}.
Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
of the form ``$c$ inherited from $p$ on the genomic segment from $x$ to $y$'',
where $c$ and $p$ are (haploid) genomes, and $x$ and $y$ are genetic coordinates.
We record each such statement in an ``edge'',
often summarized as the tuple $(p,c,x,y)$.
By ``describes the history'',
we mean that each node in the ARG represents a specific (ancestral) genome,
and that the ARG specifies the genealogical tree describing how the samples are related
at each point on the genome.

@@

Crucial for the equivalence between the two representations is the presence of
nodes that only have a single child along one or more local trees --
i.e., are \emph{locally unary}~\citep{wong_general_2023}.
Recombinant nodes (in blue) have a single child in all trees.
However, common ancestry nodes can be locally unary as well:
for instance, sample $d$ inherited a longer stretch of genome from ancestor $f$
than sample $c$.
In the top plot, this is visible because the pink bar below node $f$
extends further to the left than does the blue bar,
while in the bottom plot, this is visible because node $f$ is parent to $d$
in the second tree, but not to node $c$, and so is unary.

@@

These locally unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify %~\citep{wong_general_2023,kelleher_efficient_2018} 
the ARG by removing all recombinant nodes
($A$, $B$, and $C$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombinant nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombinant nodes we no longer know the precise time of each recombination event,
but we do still have constraints on it;
for instance, we know that $C$ happened somewhere between node $f$ and node $h$.
More generally, a recombinant node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, Fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG inference methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently,
\argweaver{} \citep{rasmussen_genome-wide_2014},
\tsinfer{} \citep{kelleher_inferring_2019},
\kwarg{} \citep{ignatieva_kwarg_2021}, and
\texttt{SINGER} \citep{deng_robust_2024} estimate ARGs 
including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation, kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).
See~\cite{fritze2024forest} for more discussion on unary nodes
and haplotype aware ARG comparison metrics.
